---
title: "Chess Tournament Elo Expected Score Analysis"
author: "Mark Hamer"
date: "2026-03-02"
output: html_document
---

## Overview

This project looks at whether the chess players in our tournament actually 
performed the way their ratings suggested they should. Using the Elo formula, 
I calculated what score each player was *expected* to get based on who they 
played, then compared that to what they actually scored. The players with the 
biggest positive gap overperformed — they beat the odds. The ones with the 
biggest negative gap underperformed — their rating said they should have done 
better.

---

## The Elo Formula

The expected score for Player A against a single opponent Player B is:

$$E = \frac{1}{1 + 10^{(R_B - R_A) / 400}}$$

So if two players have the same rating, each expects 0.5 — a coin flip. A 
400 point gap means the stronger player wins about 90% of the time. I add up 
the expected scores across all 7 rounds to get each player's total expected 
score for the tournament. Byes and unplayed rounds get skipped since there's 
no opponent to calculate against.

*Formula source: https://www.chess.com/terms/elo-rating-chess*

---

## Step 1: Load Libraries and Parse Tournament Data

```{r setup, message=FALSE, warning=FALSE}
library(stringr)
library(dplyr)

# Re-read and parse the tournament file (same as Project 1)
raw_lines <- readLines("tournamentinfo.txt", warn = FALSE)
player_lines <- raw_lines[!grepl("^-", raw_lines)]
player_lines <- player_lines[3:length(player_lines)]

row1 <- player_lines[seq(1, length(player_lines), by = 2)]
row2 <- player_lines[seq(2, length(player_lines), by = 2)]

# Extract fields
player_name <- str_trim(str_extract(row1, "(?<=\\|)[^|]+(?=\\|)"))
total_pts   <- as.numeric(str_trim(str_extract(row1, "(?<=\\|)\\s*\\d+\\.\\d+\\s*(?=\\|)")))
pre_rating  <- as.numeric(str_extract(str_extract(row2, "R:\\s*\\d+"), "\\d+"))
pair_nums   <- as.numeric(str_trim(str_extract(row1, "^\\s*\\d+")))

# Build lookup and opponents list
rating_lookup <- setNames(pre_rating, pair_nums)

get_opponents <- function(line) {
  as.numeric(na.omit(str_trim(str_extract_all(line, "(?<=[WLDX])\\s*(\\d+)")[[1]])))
}
opponents <- lapply(row1, get_opponents)
```

---

## Step 2: Calculate Elo Expected Score

```{r elo, message=FALSE, warning=FALSE}
# Elo expected score formula
elo_expected <- function(rating_a, rating_b) {
  1 / (1 + 10^((rating_b - rating_a) / 400))
}

# Calculate expected score for each player across all their games
expected_score <- sapply(seq_along(opponents), function(i) {
  opp_ids <- opponents[[i]]
  opp_ratings <- rating_lookup[as.character(opp_ids)]
  sum(sapply(opp_ratings, function(r) elo_expected(pre_rating[i], r)))
})

expected_score <- round(expected_score, 2)

head(data.frame(player_name, total_pts, expected_score))
```

---

## Step 3: Compare Actual vs Expected and Find Top 5

```{r results, message=FALSE, warning=FALSE}
# Build the comparison data frame
elo_df <- data.frame(
  Player_Name    = player_name,
  Actual_Score   = total_pts,
  Expected_Score = expected_score,
  Difference     = round(total_pts - expected_score, 2)
)

# Top 5 overperformers
overperformers <- elo_df %>% arrange(desc(Difference)) %>% head(5)

# Top 5 underperformers
underperformers <- elo_df %>% arrange(Difference) %>% head(5)

cat("TOP 5 OVERPERFORMERS\n")
print(overperformers)

cat("\nTOP 5 UNDERPERFORMERS\n")
print(underperformers)
```

---

## Discussion

Aditya Bajaj was the biggest surprise — rated 1384 going in but scoring a 
perfect 6 points against much stronger players. The formula only expected him 
to score about 2 points, so he overperformed by 4. Jacob Alexander Lavalley 
is even wilder — the math basically gave him a 0% chance but he still managed 
3 points.

On the flip side, Loren Schwiebert had the highest rating of any underperformer 
at 1745 but only scored 3.5 — almost 3 points below what his rating predicted. 
That's a rough tournament.

The takeaway is that Elo tells you probabilities, not results. Seven rounds 
isn't enough to iron out the variance, which is exactly why upsets happen.
